---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "65%",
  fig.align = "center"
)
```

## Spatial Censored Regression Models in R: The ```CensSpatial``` Package

<div style="text-align: justify"> The ```CensSpatial``` package allows the analysis of spatial censored observations. It provides  tools for simulation, estimation and prediction of Gaussian processes with spatially correlated and censored data. Four estimation procedures can be found within the package, the Seminaive method proposed in  Schelin and
Sjöstedt-de Luna (2014), the stochastic approximation (SAEM) proposed by Ordonez et al. (2018) and two naive approximations (Naive 1 and Naive 2), which impute
censored observations by using the limit of detection ($LOD$) and $\frac{LOD}{2}$ respectively. 

Next, we will illustrate the package functionalities by using the the rainfall dataset
analyzed in  Diggle and Ribeiro Jr (2002). Since this dataset does not contain
censored data, we generated it artificially for the first 100 observations, in this case,
we consider a level of censoring of 25%. The code below, shows this procedure 
</div>


```{r code1}
library(geoR)
 data(parana)

### training dataset
dataobs = data.frame(parana$coords[1:100,],parana$data[1:100])

#### test dataset 
datapred = data.frame(parana$coords[101:143,],parana$data[101:143])

### manageable names 
names(dataobs) = c('c1','c2','y')
names(datapred) = names(dataobs)

######### generating censored observations ###########

lod = quantile(dataobs$y,probs = 0.25) ### limit of detection (LOD)
dataobs$y[dataobs$y<=lod] = lod
cc = as.numeric(dataobs$y == lod)
dataobs$cc = cc
```


<div style="text-align: justify"> In order to specify the mean and covariance components of the model, we use some descriptive graphs. Figure
1 depicts the rainfall dataset, where the left panel shows the average rainfall in each coordinate and the right panel
shows the sample variogram for the observed data considering a Gaussian covariance structure. From this figure, we
can conclude that it is plausible to assume a linear trend of the rainfall with respect to its coordinates. The behavior
showed in the sample variogram seems to indicate that it is also plausible to consider a Gaussian covariance structure
for the stochastic component. </div>

```{r,fig.cap = "sdasds", echo=FALSE}
library(scatterplot3d)
datare = dataobs[,-4]
datare = as.geodata(datare)
par(mfrow=c(1,2))
adj2=variofit(variog(datare),cov.model="gaussian",messages = FALSE)
scatterplot3d(x=dataobs[,1],y=dataobs[,2],z=dataobs[,3],pch=20,box=F, color = "gray",xlab= "East", ylab = "North", zlab = "Average rainfall")
plot(variog(datare,breaks=seq(0,320,length=10)),pch=20)
lines.variomodel(seq(0,400,length=14),cov.model = "gaussian", cov.pars =adj2$cov.pars, nugget = 299.53)

```



```{r,fig.cap = "sdasds", echo=FALSE}
cov.ini=c(200,700); y=dataobs$y; cc=dataobs$cc
coords=dataobs[,1:2]; coordspred=datapred[,1:2]
```

### Estimation and prediction utilities

<div style="text-align: justify">
Now, we proceed to fit the selected model through the available algorithms within the package. The code below show the sintax for each method, arguments related to data structure, model specification and convergence items such as initial values must be specified.
</div>

```{r}
library(CensSpatial)
set.seed(123)


### SAEM sintax

SAEMres = SAEMSCL(cc,y,cens.type="left",trend="1st",coords=coords,
               kappa=2, cov.model="gaussian", fix.nugget=F,nugget=170,
               inits.sigmae=68,inits.phi=900, search=T,
               lower=c(0.0001,0.0001),upper=c(50000,5000),
               M=15,perc=0.25,MaxIter=500,pc=0.2)


### Seminaive sintax
seminaiveres = Seminaive(data=dataobs,y.col=3,coords.col=1:2,
                         covar=F,cc=as.matrix(cc),trend="1st",
                         cov.model="gaussian",fix.nugget=F,
                         nugget=170,kappa=2,
                         thetaini=c(700,200),
                         cons=c(0.1,2,0.5),MaxIter=500,
                         copred=coordspred, cutoff = lod)


### Naive 1 and Naive 2 sintax
naive12res = algnaive12(data=dataobs,y.col=3,coords.col=1:2,covar=F,
                        cc=as.matrix(cc),trend="1st",
                        cov.model="gaussian",fix.nugget=F,nugget=170,kappa=2,
                        thetaini=c(700,200),copred=coordspred,
                        cutoff=lod)

```

<div style="text-align: justify"> The function summary allows to print summarized results for each algorithm. The output below shows the use of this function for the SAEM algorithm. </div>


```{r}
summary(SAEMres)

```

<div style="text-align: justify">  Since the output of each algorithm is a list, criteria such as the log-likelihood, the Akaike and Bayesian information (AICand BIC respectively), can be recovered through the instructions ```obj$loglik```, ```obj$AIC``` and ```obj$BIC```, where obj is an object generated by any of the available
functions for estimation. It can be seen that the SAEM algorithm shows better results for all the criteria in comparison with the remaining methods.</div>


```{r}
##### Results #####

saemres = round(c(SAEMres$theta, SAEMres$loglik, SAEMres$AIC, SAEMres$BIC),3)
Seminaiveres = round(c(seminaiveres$theta, seminaiveres$loglik, seminaiveres$AIC, seminaiveres$BIC),3)
naive1res = round(c(naive12res$theta1, naive12res$loglik1, naive12res$AIC1,naive12res$BIC1),3)
naive2res = round(c(naive12res$theta2, naive12res$loglik2, naive12res$AIC2,naive12res$BIC2), 3)


tableres = rbind(Seminaiveres,naive1res,naive2res,saemres )
colnames(tableres) = c(paste0("beta",0:2), "sigma2","phi","tau2","loglik","AIC", "BIC")
rownames(tableres) = c("Seminaive","Naive_1","Naive_2","SAEM (best)")

tableres 
```

<div style="text-align: justify">  Regarding predictions,  ```CensSpatial``` allows to generate intensity plots of both predictions and their standard deviations considering the SAEM algorithm through the function  ```predgraphics```.  As main arguments we can find ```grid1```,
which consists of a grid of coordinates to construct the plot for prediction, ```xpred```,
which are the covariate values defined for this grid, ```est```, an object of the class
```SAEMSpatialCens```, and ```sdgraph```, a logical argument which is ```TRUE``` by
default, providing the standard deviation of the predictors. </div>

```{r}
coorgra1=seq(min(coords[,1]),max(coords[,1]),length=50)
coorgra2=seq(min(coords[,2]),max(coords[,2]),length=50)
grid1=expand.grid(x=coorgra1,y=coorgra2)

xpred=cbind(1,grid1)

graf=predgraphics(xpred=xpred,est=SAEMres,grid1=grid1,points=T,sdgraph=FALSE,
                  colors = gray.colors(300),obspoints = sample(1:sum(cc==0),50),main2="")
```

<div style="text-align: justify">
Finally, the function ```localinfmeas``` allows to perform local influence diagnostics by considering three schemes, perturbation in the response variable, in the scale matrix and in the explanatory covariates. In this case, we run all the schemes by specifying the option ```type.plot = all```. The code below shows the output of this function, as it can be seen, atypical observations as well as diagnostic plots can be obtained from this tool.

</div>




```{r}
dp = localinfmeas(est = SAEMres,fix.nugget= FALSE,diag.plot=TRUE,type.plot="all",c=3)

### explanatory variables perturbation ##
dp$expvper

### scale matrix perturbation##
dp$smper
```


## References


Diggle, P. (2002). Analysis of longitudinal data. Oxford University Press,
USA.
6

Ordonez, J. A., D. Bandyopadhyay, V. H. Lachos, and C. R. Cabral (2018).
Geostatistical estimation and prediction for censored responses. Spatial
statistics 23, 109–123

Ordonez, J., C. Galarza, and V. Lachos (2022). CensSpatial:
Censored spatial models. R package version 3.5, URL
\url{http://CRAN.R-project.org/package=CensSpatial}

Schelin, L. and S. Sjostedt-de Luna (2014). Spatial prediction in the presence
of left-censoring. Computational Statistics and Data Analysis 74.
